---
title: "Lab2 - AN - Analysis"
author: "Alec Naidoo"
date: "2024-03-22"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(wooldridge)
library(car)
library(lmtest)
library(sandwich)
library(stargazer)
input_data <- read.csv("/Users/alecnaidoo/Downloads/asecpub23csv/pppub23.csv")
```

# Intro
So far, my goal is to find any covariate trends with variables and age, in order to describe welfare/government assistance programs. We initialize the analysis with a basic model, and then add complexity with covariate measures that we find in the data.

First we can perform data wrangling. Ultimately we are only left with 734 of 146k observations (0.5% of dataset), that the survey reports have received public assistance.

Second, by individual plotting we can begin to understand relationships of age with other subtopics related to non-cash benefits and supplementary poverty measures, answered through the survey results.

# Data Wrangling

```{r data wrangling, include = FALSE}
df <- input_data %>%
  mutate(unique_id = paste0(PH_SEQ, PPPOS))

# check if records are unique
if (anyDuplicated(df$unique_id) == 0) {
  message("No duplicates found for unique_id")
} else {
  message("duplicates found in unique_id")
}

person_df <- df %>%
  filter(!A_AGE %in% c(85, 80) # Remove categorical data for those grouped in 80 and 85
         & PAW_YN == 1 # Set to the universe of individuals that received CASH assistance from a state or county welfare program
         & PAW_VAL > 0) 


# if ("PAW_VAL" %in% names(person_df)) {
#   print("Variable exists!")
# } else {
#   print("Variable does not exist.")
# }
```

```{r data wrangling, include = FALSE}
new <- df %>%
  filter(!A_AGE %in% c(85, 80) # Remove categorical data for those grouped in 80 and 85
         & OI_YN == 1 ) 

```



# Build Basic Model
This will serve as our baseline when comparing more complex models.
```{r basic model creation, include = TRUE}

model <- lm(PAW_VAL ~ A_AGE, data = person_df)

print(summary(model))
```

## Basic Plot of Age versus Govt Assistance Income
```{r pressure, echo=TRUE}
# Create ggplot
ggplot(person_df, aes(x = A_AGE, y = PAW_VAL)) +
  geom_point() +  # Add points
  labs(x = "Age", y = "PAW_VAL", title = "PAW_VAL by Age") + # Add axis labels and title
  geom_smooth(method = "lm", se = FALSE, color = "red") 
```

## Scatter Plot all relationships withing subtopics "Non-Cash Benefits" and "Supplementary Poverty Measures" (can include more vars)
``` {r adding tons of welfare parameters to look for relationships}

# Create scatterplot matrix focusing on the relationship with A_AGE
pairs(~ A_AGE + PAW_VAL + PAW_MON + PAW_TYP + PAW_YN + PENPLAN + WICYN + CHCARE_YN + CHELSEW_YN + CHSP_VAL + CHSP_YN + CSP_VAL + CSP_YN, data = person_df)


```
Although this accomplishes the goal, most of the data is unreadable and hence we proceed with the below method to find relationships.

## Cycle through individual plots of Age and 'Other' Variables with regression line (can include more vars)
``` {r relationship between csp_value, echo=TRUE}

# List to store plots
plots_list <- list()

# List of variables
variables_to_plot <- c("PAW_VAL", "PAW_MON", "PAW_TYP", "PAW_YN", "PENPLAN", "WICYN", "CHCARE_YN", "CHELSEW_YN", "CHSP_VAL", "CHSP_YN", "CSP_VAL", "CSP_YN")

# Loop through each variable and create scatterplot against A_AGE
for (variable in variables_to_plot) {
  plot <- ggplot(person_df, aes_string(x = "A_AGE", y = variable)) +
    geom_point() +
    labs(x = "Age", y = variable) +
    ggtitle(paste(variable, "by Age")) +
    geom_smooth(method = "lm", se = FALSE, color = "red") 
  
  plots_list[[variable]] <- plot  # Store the plot in the list
}

# Print each plot
for (variable in names(plots_list)) {
  print(plots_list[[variable]])
}


```
## Plot Age vs Govt Assistance Income wrt observed categorical variables
While above plots are interesting, we can further examing any particular trends in the main relationship wrt indicator variables.

``` {r relationship between age and welfare_income wrt var colors, echo=TRUE}

# List to store plots
plots_list <- list()

# List of variables
categorical_variables_to_plot <- c("A_SEX","PAW_MON", "PAW_TYP", "PENPLAN", "WICYN", "CHCARE_YN", "CHELSEW_YN", "CHSP_YN", "CSP_YN", "SEX_WICYN", "PRDTRACE")

# Copy dataset in case
person_df2 <- person_df

# Create Combined Columns
person_df2$SEX_WICYN <- interaction(person_df2$A_SEX, person_df2$WICYN, drop = TRUE)

# Make variables categorical in R
person_df2$A_SEX <- factor(person_df2$A_SEX)
person_df2$PAW_MON <- factor(person_df2$PAW_MON)
person_df2$PAW_TYP <- factor(person_df2$PAW_TYP)
person_df2$PENPLAN <- factor(person_df2$PENPLAN)
person_df2$WICYN <- factor(person_df2$WICYN)
person_df2$CHCARE_YN <- factor(person_df2$CHCARE_YN)
person_df2$CHELSEW_YN <- factor(person_df2$CHELSEW_YN)
person_df2$CHSP_YN <- factor(person_df2$CHSP_YN)
person_df2$CSP_YN <- factor(person_df2$CSP_YN)

person_df2$SEX_WICYN <- factor(person_df2$SEX_WICYN)
person_df2$PRDTRACE <- factor(person_df$PRDTRACE)

# Loop through each variable and create scatterplot against A_AGE
for (variable in categorical_variables_to_plot) {
  plot <- ggplot(person_df2, aes_string(x = "A_AGE", y = "PAW_VAL", color = variable)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(x = "Age", y = variable) +
    ggtitle(paste("Assisted Income by Age wrt: ", variable))
  
  plots_list[[variable]] <- plot  # Store the plot in the list
}

# Print each plot
for (variable in names(plots_list)) {
  print(plots_list[[variable]])
}


```
## Test Log manipulation and factorizing AGE variable

``` {r}
model_a <- lm(PAW_VAL ~ A_AGE + I(A_AGE^2), data = person_df2)

summary(model_a)
```



```{r compare basic model and complex model 1, include = TRUE}

model <- lm(PAW_VAL ~ A_AGE, data = person_df)
model2 <- lm(PAW_VAL ~ log(A_AGE) + A_AGE, data = person_df2)
model3 <- lm(log(PAW_VAL) ~ A_AGE, data = person_df2)
model4 <- lm(PAW_VAL ~ as.factor(A_AGE), data = person_df2)
print(summary(model))
print(summary(model2))
print(summary(model3))
print(summary(model4))
```

We don't see much increase in explained variation by these simple manipulations, although for model 2 (applying log to the outcome variable 'PAW_VAL') we observe a statistically significant coefficient for age (**). Naive interpretation would be that age is a more consistent predictor of log(govt assistance income).

```{r plot age vs log(paw_val), echo=TRUE}
ggplot(person_df, aes(x = A_AGE, y = log(PAW_VAL))) +
  geom_point() +  # Add points
  labs(x = "Age", y = "log PAW_VAL", title = "PAW_VAL by Age") + # Add axis labels and title
  geom_smooth(method = "lm", se = FALSE, color = "red") 
```


## Include more complexity in the model
```{r compare basic model and complex model 1, include = TRUE}

model <- lm(PAW_VAL ~ A_AGE, data = person_df)
model2 <- lm(log(PAW_VAL) ~ A_AGE, data = person_df2)
model3 <- lm(log(PAW_VAL) ~ A_AGE + WICYN, data = person_df2)
model4 <- lm(log(PAW_VAL) ~ A_AGE*SEX_WICYN, data = person_df2)
print(summary(model))
print(summary(model2))
print(summary(model3))
print(summary(model4))
```


```{r sandbox graphs, echo=TRUE}
ggplot(person_df, aes(x = A_AGE, y = log(PAW_VAL))) +
  geom_point() +  # Add points
  labs(x = "Age", y = "log PAW_VAL", title = "PAW_VAL by Age") + # Add axis labels and title
  geom_smooth(method = "lm", se = FALSE, color = "red") 
```






